---
title: "Get Started with llmr"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Get Started with llmr}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  eval = TRUE 
)
```

# Introduction

The **llmr** package provides a flexible R interface to large language models (LLMs) with support for multiple providers, conversation management, and tool use. This vignette will get you started with the basics: creating an agent and equipping it with a simple calculator tool.

## What You'll Learn

By the end of this guide, you'll understand how to:

- Set up an LLM provider (Anthropic's Claude)
- Create an AI agent
- Add a calculator tool to your agent
- Have a conversation where the agent can perform calculations

## Prerequisites

Before you begin, you'll need:

1. **An API key** from a supported LLM provider (we'll use Anthropic's Claude in this example)
2. **The mcpr package** installed (for creating tools)

```{r setup}
library(llmr)
library(mcpr)
```

## Step 1: Set Up Your Provider

First, you need to configure your LLM provider. We'll use Anthropic's Claude, but llmr also supports OpenAI.

```{r provider-setup}
# Create a provider instance
provider <- new_anthropic()
```

**Important:** Replace `"your-anthropic-api-key-here"` with your actual Anthropic API key. You can get one from [Anthropic's website](https://console.anthropic.com/).

## Step 2: Create Your First Agent

An agent is an AI assistant that can use tools to help solve problems. Let's create a simple calculator agent:

```{r create-agent}
# Create an agent with a descriptive name
agent <- new_agent("calculator", provider)
```

That's it! You now have a basic agent, but it can't do much yet. Let's give it some capabilities.

## Step 3: Create a Calculator Tool

Tools allow your agent to perform specific actions. We'll create a calculator tool that can perform basic arithmetic operations:

```{r create-calculator}
# Create a calculator tool
calculator <- new_tool(
  name = "calculator",
  description = "Performs basic arithmetic operations",
  input_schema = schema(
    properties = properties(
      operation = property_enum(
        "Operation",
        "Math operation to perform",
        values = c("add", "subtract", "multiply", "divide"),
        required = TRUE
      ),
      a = property_number("First number", "First operand", required = TRUE),
      b = property_number("Second number", "Second operand", required = TRUE)
    )
  ),
  handler = function(params) {
    result <- switch(
      params$operation,
      "add" = params$a + params$b,
      "subtract" = params$a - params$b,
      "multiply" = params$a * params$b,
      "divide" = params$a / params$b
    )
    response_text(result)
  }
)
```

Let's break down what this tool does:

- **Name and description**: Identifies the tool and explains its purpose
- **Input schema**: Defines what parameters the tool expects:
  - `operation`: Must be one of "add", "subtract", "multiply", or "divide"
  - `a` and `b`: The two numbers to operate on
- **Handler function**: The actual R code that performs the calculation

## Step 4: Add the Tool to Your Agent

Now we need to give our agent access to the calculator tool:

```{r add-tool}
# Add the calculator tool to the agent
agent <- add_tool(agent, calculator)
```

Your agent is now equipped with calculation capabilities!

## Step 5: Have a Conversation

Let's test our calculator agent by asking it to perform some calculations:

```{r conversation}
# Ask the agent to perform a calculation
request(agent, "What is 15 multiplied by 7?")
get_last_message(agent)
```

## How It Works

When you ask the agent a question that requires calculation:

1. **The agent analyzes your request** and determines it needs to perform math
2. **It calls the calculator tool** with the appropriate operation and numbers
3. **The tool performs the calculation** and returns the result
4. **The agent incorporates the result** into a natural language response

This happens automatically - you don't need to explicitly tell the agent to use the calculator tool.

## Step 6: Adding Multiple Tools

Our agent can currently perform calculations, but let's expand its capabilities by adding a simple text formatter tool. This will demonstrate how a single request can invoke multiple tools when needed.

```{r create-formatter}
# Create a text formatter tool
formatter <- new_tool(
  name = "convert",
  description = "Convert a number to Roman numerals",
  input_schema = schema(
    properties = properties(
      value = property_number(
        "Number",
        "The Number to convert",
        required = TRUE
      )
    )
  ),
  handler = function(params) {
    response_text(as.roman(params$value))
  }
)

# Add the formatter tool to our agent
agent <- add_tool(agent, formatter)
```

Now our agent has two tools at its disposal: a calculator and a text formatter. Let's ask it a question that requires both tools:

```{r multi-tool-request}
# Ask a question that requires both calculation and text formatting
request(agent, "Calculate 25 times 4, and convert the result to Roman numerals.")
get_last_message(agent)
```

## How Multiple Tool Use Works

When you send a request that requires multiple tools:

1. **The agent analyzes your request** and determines which tools it needs
2. **It can call multiple tools in sequence** within a single request:
   - First, it uses the calculator to compute 25 × 4 = 100
   - Then, it uses the formatter to convert "C" (Roman numeral for 100) to uppercase
3. **The agent combines all the results** into a coherent response

The beauty of this approach is that you don't need to manage the tools manually - the agent decides which tools to use and in what order based on the task at hand.

## What's Next?

Now that you understand the basics, you can:

- **Explore other providers**: Try using OpenAI instead of Anthropic
- **Create custom tools**: Build tools for your specific use cases
- **Combine multiple tools**: Give your agent access to several different capabilities
- **Use workflows**: Chain multiple agents together for complex tasks

## Key Concepts Summary

- **Provider**: The LLM service (Anthropic, OpenAI, etc.)
- **Agent**: An AI assistant that can use tools
- **Tool**: A function that extends your agent's capabilities
- **Schema**: Defines the structure of tool inputs
- **Handler**: The R function that actually performs the tool's work
- **Multiple Tool Use**: The ability for an agent to use several tools in a single request

The llmr package makes it easy to create powerful AI agents that can interact with your R environment and perform real work. Start simple with tools like our calculator and formatter, then gradually build more sophisticated capabilities as you become comfortable with the framework.
