---
title: "MCP Integration: From Development to Production"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{MCP Integration: From Development to Production}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  eval = FALSE
)
```

# Introduction

The **Model Context Protocol (MCP)** is a standardized way to connect AI assistants with external tools and data sources. The **llmr** package provides seamless integration with MCP through the **mcpr** package, enabling a powerful "lift and shift" approach to tool development.

This vignette demonstrates how to take the calculator tool from the "Get Started" guide and deploy it as a production-ready MCP server, while maintaining the same simple client interface.

## What You'll Learn

By the end of this guide, you'll understand how to:

- Transform inline tools into standalone MCP servers
- Connect llmr agents to external MCP services
- Leverage the "lift and shift" approach for production deployment
- Scale your tool architecture from development to production

## The Development-to-Production Journey

### Development Phase: Inline Tools
In the "Get Started" vignette, we created a calculator tool directly within our agent:

```{r inline-tool}
# Development approach - tool embedded in agent
calculator <- new_tool(
  name = "calculator",
  description = "Performs basic arithmetic operations",
  input_schema = schema(
    properties = properties(
      operation = property_enum(
        "Operation",
        "Math operation to perform",
        values = c("add", "subtract", "multiply", "divide"),
        required = TRUE
      ),
      a = property_number("First number", "First operand", required = TRUE),
      b = property_number("Second number", "Second operand", required = TRUE)
    )
  ),
  handler = function(params) {
    result <- switch(
      params$operation,
      "add" = params$a + params$b,
      "subtract" = params$a - params$b,
      "multiply" = params$a * params$b,
      "divide" = params$a / params$b
    )
    response_text(result)
  }
)
agent <- add_tool(agent, calculator)
```

This approach is perfect for:
- **Rapid prototyping**
- **Testing new functionality**
- **Simple, single-use tools**

### Production Phase: MCP Servers
For production deployment, we can "lift and shift" the same tool logic into an MCP server:

```{r mcp-approach}
# Production approach - tool as external MCP server
client <- new_client_io(command = "Rscript", args = "calculator-server.R")
agent <- register_mcp(agent, client)
```

This approach provides:
- **Scalability** - Tools run as separate processes
- **Reusability** - Multiple agents can share the same tools
- **Maintainability** - Tools can be updated independently
- **Security** - Tools run in isolated environments

## Step 1: Understanding the MCP Server

Let's transform our calculator tool into a standalone MCP server. Here's what the server code looks like:

```{r server-code}
library(mcpr)

# Create the same calculator tool from the Get Started guide
calculator <- new_tool(
  name = "calculator",
  description = "Performs basic arithmetic operations",
  input_schema = schema(
    properties = properties(
      operation = property_enum(
        "Operation",
        "Math operation to perform",
        values = c("add", "subtract", "multiply", "divide"),
        required = TRUE
      ),
      a = property_number("First number", "First operand", required = TRUE),
      b = property_number("Second number", "Second operand", required = TRUE)
    )
  ),
  handler = function(params) {
    result <- switch(
      params$operation,
      "add" = params$a + params$b,
      "subtract" = params$a - params$b,
      "multiply" = params$a * params$b,
      "divide" = params$a / params$b
    )
    response_text(result)
  }
)

# Create an MCP server and add the calculator tool
mcp_server <- new_server(
  name = "Calculator Server",
  description = "A production-ready calculator service",
  version = "1.0.0"
)

# Add the tool to the server
mcp_server <- add_capability(mcp_server, calculator)

# Start the server (listening on stdin/stdout)
serve_io(mcp_server)
```

**Key Insight**: Notice how the tool definition is **identical** to the inline version. This is the power of the "lift and shift" approach - no code changes required!

The llmr package includes this exact server as a ready-to-use example:

## Step 2: Create the MCP Client

Now let's create an agent that connects to our MCP server:

```{r client-setup, eval=TRUE}
library(llmr)
library(mcpr)

# Set up your provider (same as before)
provider <- new_anthropic()

# Create an agent
agent <- new_agent("calculator", provider)
```

## Step 3: Connect to the MCP Server

Instead of adding tools directly, we'll connect to our external MCP server. The llmr package includes a ready-to-use calculator server:

```{r mcp-connection, eval=TRUE}
# Get the path to the example calculator server included with llmr
server_path <- system.file("examples", "calculator-server.R", package = "llmr")

# Create an MCP client that connects to our calculator server
calculator_client <- new_client_io(
  command = "Rscript",
  args = server_path,
  name = "calculator"
)

# Register the MCP client with our agent
agent <- register_mcp(agent, calculator_client)
```

**What's happening here:**
- `new_client_io()` creates a client that launches our server as a subprocess
- The server communicates via stdin/stdout (standard MCP protocol)
- `register_mcp()` makes all server tools available to the agent

## Step 4: Use the Agent (Same Interface!)

The beauty of this approach is that the client interface remains **exactly the same**:

```{r usage-examples, eval=TRUE}
# Same usage as the inline tool version
request(agent, "What is 25 multiplied by 4?")
get_last_message(agent)

# Complex calculations work the same way
request(agent, "I have 150 dollars, spend 47.50 on dinner and 23.75 on a book. How much is left?")
get_last_message(agent)
```

## The "Lift and Shift" Advantage

### Development Workflow
1. **Start with inline tools** for rapid prototyping
2. **Test and refine** your tool logic
3. **Lift the tool definition** to an MCP server
4. **Shift your agent** to use the MCP client
5. **Deploy to production** with no client code changes

### Benefits of This Approach

#### ðŸš€ **Scalability**
```{r scalability-example}
# Multiple agents can share the same calculator service
agent1 <- new_agent("financial_advisor", provider)
agent2 <- new_agent("math_tutor", provider)
agent3 <- new_agent("data_analyst", provider)

# All connect to the same MCP server
register_mcp(agent1, calculator_client)
register_mcp(agent2, calculator_client)
register_mcp(agent3, calculator_client)
```

#### ðŸ”„ **Reusability**
- One calculator server serves multiple use cases
- Tools become organizational assets, not agent-specific code
- Easy to discover and share tools across teams

#### ðŸ›¡ï¸ **Security & Isolation**
- Tools run in separate processes with controlled permissions
- Server crashes don't affect the main agent
- Easy to implement resource limits and monitoring

#### ðŸ”§ **Maintainability**
- Update tool logic without touching agent code
- Version control tools independently
- Deploy tool updates without agent restarts

## Advanced MCP Patterns

### Multiple MCP Services
You can connect to multiple MCP servers for different capabilities:

```{r multiple-services}
# Connect to different specialized services
calculator_client <- new_client_io(command = "Rscript", args = "calculator-server.R")
weather_client <- new_client_io(command = "Rscript", args = "weather-server.R")
database_client <- new_client_io(command = "Rscript", args = "database-server.R")

# Register all services with the agent
agent <- register_mcp(agent, calculator_client)
agent <- register_mcp(agent, weather_client)  
agent <- register_mcp(agent, database_client)
```

### Hybrid Approach
You can mix inline tools with MCP services:

```{r hybrid-approach}
# Quick inline tool for simple tasks
simple_tool <- new_tool(name = "timestamp", ...)
agent <- add_tool(agent, simple_tool)

# External MCP service for complex operations
complex_client <- new_client_io(...)
agent <- register_mcp(agent, complex_client)
```

## Production Deployment Considerations

### Server Management
- Use process managers (systemd, Docker, etc.) for MCP servers
- Implement health checks and automatic restarts
- Monitor server performance and resource usage

### Security
- Run MCP servers with minimal required permissions
- Use network isolation where appropriate
- Implement authentication for sensitive tools

### Scaling
- Deploy MCP servers on separate machines for heavy workloads
- Use load balancers for high-availability tool services
- Implement caching for frequently-used operations

## Migration Strategy

### Phase 1: Development
```{r phase1}
# Start with inline tools
agent <- add_tool(agent, my_tool)
```

### Phase 2: Testing
```{r phase2}
# Test MCP version alongside inline version
agent <- add_tool(agent, my_tool)  # Keep original
agent <- register_mcp(agent, mcp_client)  # Add MCP version
```

### Phase 3: Production
```{r phase3}
# Remove inline tool, use only MCP
agent <- register_mcp(agent, mcp_client)
```

## Summary

The MCP integration in llmr provides a seamless path from development to production:

1. **Develop quickly** with inline tools
2. **Lift tool definitions** to MCP servers without code changes
3. **Shift agents** to use MCP clients with minimal changes
4. **Scale confidently** with production-ready architecture

This approach gives you the best of both worlds: the speed of inline development and the robustness of distributed production systems.

### Key Takeaways
- **Same tool logic** works in both inline and MCP modes
- **Client interface remains unchanged** during migration
- **MCP provides production benefits** without development complexity
- **Gradual migration** is possible and recommended

The Model Context Protocol integration makes llmr a powerful platform for building scalable AI agent systems that can grow from prototype to production seamlessly.
