---
title: "Workflow Basics: Linear Chains"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Workflow Basics: Linear Chains}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  eval = TRUE
)
```

# Introduction

Workflows in **llmr** allow you to chain together multiple processing steps to create sophisticated data processing pipelines. The most fundamental workflow pattern is the **linear chain**, where data flows sequentially through a series of steps.

This vignette introduces the basics of workflows by showing how to create simple linear chains using both regular R functions and AI agents as steps.

## What You'll Learn

By the end of this guide, you'll understand how to:

- Create workflow steps from R functions
- Create workflow steps from AI agents  
- Chain steps together using the `%->%` operator
- Execute workflows with input data
- Understand the flexibility of mixing functions and agents

## Prerequisites

```{r setup}
library(llmr)
library(ellmer)
```

## Basic Workflow Concepts

### Steps
A **step** is the fundamental building block of a workflow. Steps can be created from:
- **R functions** - For deterministic processing
- **AI agents** - For intelligent, context-aware processing

### Chains
A **chain** connects steps together using the `%->%` operator, creating a pipeline where the output of one step becomes the input of the next.

### Execution
**Execution** runs the entire workflow with your input data, passing results through each step in sequence.

## Example 1: Function-Based Linear Workflow

Let's start with a simple workflow using regular R functions:

```{r function-workflow}
# Define simple processing functions
add_ten <- function(x) {
  cat("Adding 10 to", x, "\n")
  x + 10
}

multiply_by_two <- function(x) {
  cat("Multiplying", x, "by 2\n")
  x * 2
}

subtract_five <- function(x) {
  cat("Subtracting 5 from", x, "\n")
  x - 5
}

# Create workflow steps and chain them together inline
math_workflow <- step(add_ten, name = "add_ten") %->%
  step(multiply_by_two, name = "multiply_two") %->%
  step(subtract_five, name = "subtract_five")

# View the workflow structure
print(math_workflow)

# Execute the workflow
input_value <- 5
result <- execute(math_workflow, input_value)
print(result)  # Should be 25: ((5 + 10) * 2) - 5 = 25
```

**Key Points:**
- Each function takes one input and returns one output
- The `step()` function wraps your function for use in workflows
- The `%->%` operator chains steps left-to-right
- `execute()` runs the entire workflow

## Example 2: Agent-Based Linear Workflow

Now let's create a workflow using AI agents for text processing:

```{r agent-workflow, eval = FALSE}
# Create specialized agents using ellmer chat objects
summarizer <- new_agent("summarizer", ellmer::chat_anthropic())
analyzer <- new_agent("analyzer", ellmer::chat_anthropic())
formatter <- new_agent("formatter", ellmer::chat_anthropic())

# Create specialized agents with specific system prompts
summarizer <- new_agent("summarizer", ellmer::chat_anthropic(
  system_prompt = "You are a text summarizer. Provide concise, one-sentence summaries."
))
analyzer <- new_agent("analyzer", ellmer::chat_anthropic(
  system_prompt = "You are a sentiment analyzer. Analyze the emotional tone and return just the sentiment (positive/negative/neutral)."
))
formatter <- new_agent("formatter", ellmer::chat_anthropic(
  system_prompt = "You are a formatter. Take the previous analysis and format it as: 'Summary: [summary] | Sentiment: [sentiment]'"
))

# Create workflow by chaining agent steps inline
text_workflow <- step(summarizer, name = "summarize") %->%
  step(analyzer, name = "analyze_sentiment") %->%
  step(formatter, name = "format_output")

# View the workflow structure
print(text_workflow)

# Execute the workflow
input_text <- "The new product launch was incredibly successful, exceeding all our expectations.
  Customers are thrilled with the innovative features and the team is celebrating this major milestone."

result <- execute(text_workflow, input_text)
print(result)
```

**Key Points:**
- Agents work just like functions in workflows
- Each agent processes the output from the previous step
- System prompts define each agent's specific role
- The workflow creates a sophisticated text processing pipeline

## Example 3: Mixed Function and Agent Workflow

The real power comes from mixing functions and agents in the same workflow:

```{r mixed-workflow, eval = FALSE}
# Function to preprocess text
clean_text <- function(text) {
  # Remove extra whitespace and convert to lowercase
  cleaned <- trimws(tolower(text))
  cat("Cleaned text:", substr(cleaned, 1, 50), "...\n")
  cleaned
}

# Function to add metadata
add_metadata <- function(analysis) {
  # Add timestamp and word count to the analysis
  timestamp <- Sys.time()
  word_count <- length(strsplit(analysis, "\\s+")[[1]])
  
  result <- paste0(
    "Analysis: ", analysis, "\n",
    "Generated: ", timestamp, "\n",
    "Word count: ", word_count
  )
  
  cat("Added metadata\n")
  result
}

# Create an agent for the core analysis
analyzer <- new_agent("text_analyzer", ellmer::chat_anthropic(
  system_prompt = "Analyze the given text and provide insights about its main themes, tone, and key messages. Be concise but thorough."
))

# Create mixed workflow: function -> agent -> function
mixed_workflow <- step(clean_text, name = "preprocess") %->%
  step(analyzer, name = "analyze") %->%
  step(add_metadata, name = "finalize")

# View the workflow structure
print(mixed_workflow)

# Execute the mixed workflow
input_text <- "   THE QUARTERLY RESULTS show STRONG GROWTH across all sectors!   "
result <- execute(mixed_workflow, input_text)
print(result)
```

**Key Points:**
- Functions and agents can be mixed freely in workflows
- Functions handle deterministic processing (cleaning, formatting)
- Agents handle intelligent analysis and interpretation
- This creates powerful hybrid processing pipelines

## Understanding Workflow Execution

### Data Flow
```{r data-flow-example, eval=FALSE}
# Input -> Step 1 -> Step 2 -> Step 3 -> Output
#   5   ->   15   ->   30   ->   25   -> Final Result
```

### Step Naming
```{r step-naming, eval = FALSE}
# Steps can be named for clarity
my_workflow <- step(my_function, name = "descriptive_name") %->%
  step(another_function, name = "another_step")

# Names appear in workflow output and debugging
print(my_workflow)  # Shows step names and visual diagram
```

### Error Handling
If any step fails, the entire workflow stops:

```{r error-handling}
# Function that might fail
risky_function <- function(x) {
  if (x < 0) stop("Cannot process negative numbers")
  x * 2
}

# Workflow will stop if risky_function fails
risky_workflow <- step(risky_function, name = "risky_step")
```

## Best Practices

### 1. Keep Steps Simple
Each step should have a single, clear responsibility:

```{r simple-steps}
# Good: Single responsibility
validate_input <- function(x) {
  if (!is.numeric(x)) stop("Input must be numeric")
  x
}

# Good: Single transformation
double_value <- function(x) x * 2

# Better than one complex function doing both
```

### 2. Use Descriptive Names
```{r descriptive-names, eval = FALSE}
# Good: Clear what each step does
workflow <- step(validate_input, name = "validate") %->%
  step(double_value, name = "double") %->%
  step(format_output, name = "format")
```

### 3. Design for Reusability
```{r reusability, eval = FALSE}
# Create reusable steps
validation_step <- step(validate_input, name = "validate")
doubling_step <- step(double_value, name = "double")

# Use in multiple workflows
workflow_a <- validation_step %->% doubling_step
workflow_b <- validation_step %->% other_step %->% doubling_step
```

## Common Patterns

### Preprocessing Pipeline
```{r preprocessing-pattern, eval=FALSE}
preprocessing <- step(validate_data, name = "validate") %->%
  step(clean_data, name = "clean") %->%
  step(normalize_data, name = "normalize")
```

### Analysis Pipeline
```{r analysis-pattern, eval=FALSE}
analysis <- step(analyzer_agent, name = "analyze") %->%
  step(extract_insights, name = "extract") %->%
  step(format_results, name = "format")
```

### Full Processing Pipeline
```{r full-pipeline, eval=FALSE}
complete_pipeline <- preprocessing %->% analysis
```

## What's Next?

Now that you understand linear workflows, you can explore:

- **Conditional workflows** - Branching based on data or conditions
- **Parallel workflows** - Processing multiple paths simultaneously  
- **Complex workflows** - Combining linear, conditional, and parallel patterns
- **Workflow composition** - Building larger workflows from smaller ones

## Summary

Linear workflows provide a powerful way to chain together processing steps:

- **Steps** can be created from both functions and agents
- **Chains** use the `%->%` operator to connect steps
- **Mixed workflows** combine the deterministic power of functions with the intelligence of agents
- **Simple patterns** can be composed into sophisticated processing pipelines

The flexibility to mix functions and agents makes llmr workflows incredibly versatile for building everything from simple data processing pipelines to complex AI-powered analysis systems.
